An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-03-29 EconGrowthSeminar.Rnw
Many minor edits. Now including better understanding of LINEX model.

% The equation for a linex model
\newcommand{\linexEqn}{$y = e \: \mathrm{e}^{\left[a_0 \left(2 - \frac{l + e}{k} \right) + a_0 c_t \left(\frac{l}{e} - 1 \right)\right]}$}

@@

\begin{frame}{Oil and Recessions}{\citep{Hamilton:2011fo}} %----------------------------
  \begin{center}
    10 of last 11 US recessions preceeded by oil price shocks
  \end{center}
\end{frame}

@@

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
    \vspace{\baselineskip}
    \pause

    Extra credit: reflect dynamics of energy cost share and prices spikes.
    \vspace{\baselineskip}
    \pause

    \footnotesize
    No known rigorous, statistical comparative studies of economic growth models for various types of economies. 
    \normalsize
  \end{center}
\end{frame}

@@

\begin{frame}{Economic Definitions} %----------------------------
  \begin{itemize}
    \item{Output ($Y$): the real market value of the total amount of produced goods in an economy (GDP).}
    \item{Capital Stock ($K$): the real market value of the total amount of goods in an economy used in the production process of output (machines, tools, infrastructure, etc.).}
    \item{Labor ($L$): the number of workers in an economy that produce output. Labor can be treated as either the total number of workers in an economy or the total number of hours worked per year in an economy.}
  \end{itemize}
\end{frame}

@@

\minorSeparator{Energy Data}

\begin{frame}{Energy Definitions} %----------------------------
  \begin{itemize}
    \item{Thermal Energy ($Q$): The heat content of an energy carrier.}
    \item{Exergy ($X$): The maximum work that can be extracted from a heat carrier by a process that brings the system into equillibrium with a thermal reservoir (typically the surrounding environment).}
    \item{Useful Work ($U$): Energy that directly generates economic activity.}
  \end{itemize}
  \begin{center}Note: indexed $Q$ and $X$ are identical unless energy mix changes significantly. \end{center}
\end{frame}

@@

\begin{frame}{Sources of Energy Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{IEA} & \multirow{3}{*}{\includegraphics[scale=0.20]{figure_other/Figure_IEA.png}} & Coal, Oil, Petroleum, & \multirow{3}{*}{GJ/year} \\
                              &                          & Nat. gas, Nuclear, & \\
                              &                          & Renewables & \\
      \hline
      \multirow{4}{*}{FAOStat} & \multirow{4}{*}{\includegraphics[scale=0.07]{figure_other/Figure_FAOStat.png}} & Fuel wood,    & \multirow{4}{*}{GJ/year} \\
                              &                          & Available livestock &                                \\
                              &                          & (1980-2011: Developed; &                         \\
                              &                          & 1991-2011: Other) &                         \\
      \hline
      \multirow{2}{*}{REXS} & \multirow{2}{*}{\includegraphics[scale=0.05]{figure_other/Figure_Warr.png}} & Energy allocation factors. &  \\
                              &                          & (1980-2000: Developed)     & \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

@@

    \begin{itemize}
      \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
      \begin{itemize}
        \item{Represents the irrepressible ``progress'' of technology}
        \item{Is independent of factor of production ($f$)}
      \end{itemize}
      \item{Used for illustration purposes here}
      \item{Example: \citet{Nel:2009wq} use $f$ = $q$}
      \item{Parameters estimated by \texttt{nls} function in \texttt{R}}
    \end{itemize}
\end{frame}

@@

\begin{frame}{Parameter Estimation for Cobb-Douglas With Energy} %----------------------------
  \begin{center}\cdeEqn\end{center}
  \begin{itemize}
    \item{Objective: determine $\lambda$, $\alpha$, $\beta$, and $\gamma$ that best fits historical data}
    \item{Perform grid search over $\gamma$: Re-estimate $\lambda$, $\alpha$, and $\beta$ at each value of $\gamma$}
    \item{Respect sum constraint: $\alpha + \beta + \gamma = 1$}
    \item{Respect absolute constraints: $0 \leq \alpha, \beta, \gamma \leq 1$} (not yet)
    \item{Parameters estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}

@@

    \item{Are we sure? (Is energy statistically significant?)\\Yes, if $CI_{\gamma}$ doesn't cross 0.0.}
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item{Energy parameter ($\gamma$) is either}
      \begin{itemize}
        \item{Small (US, JP, CN, SA, IR)}
        \item{Large CI (ZA, TZ, ZM)}
        \item{Exception: UK}
      \end{itemize}
    \vspace{\baselineskip}

@@

\begin{frame}{Comments on CES Model Equation} %----------------------------
  \begin{center} \small \cesEqn \normalsize \end{center}
  \begin{itemize}
    \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
    \item{$\delta$: split between ($kl$) and ($e$)}
    \item{$\sigma \equiv \frac{1}{1+\rho}$: elasticity of substitution ($0 \leq \sigma \leq 1$)}
    \begin{itemize}
      \item{$\sigma \rightarrow 0$: ($kl$) and ($e$) are complementary}
      \item{$\sigma \rightarrow 1$: ($kl$) and ($e$) are substitutes}
    \end{itemize}
    \item{Example: \citet{Stern:2012ey}}
    \item{Parameters estimated by \texttt{cesEst} function in \texttt{R}'s \texttt{micEcon} package}
  \end{itemize}
\end{frame}

@@

    \item{Are we sure? (Is energy statistically significant?)\\Yes, if $CI_{\delta}$ doesn't cross 0.0.}
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item{Where significant, $\delta$ is close to 1.0}
    \item{Direct energy input to economy ($q$) is unimportant as a CES predictor of economic output ($y$)}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Comments on LINEX Model Equation} %----------------------------
  \begin{center} \linexEqn \end{center}
  \begin{itemize}
    \item{Derived from thermodynamic considerations}
    \item{Foundation is energy ($e$)}
    \item{No time-dependent technology term}
    \item{Ratios among $k$, $l$, and $e$ ``trim'' energy}
    \item{Example: \citet{Warr:2006hc}}
    \item{Parameters estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}

@@

    \item{Are we sure? Yes, by definition!}
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item{LINEX parameters $a_0$ and $c_t$ are significant (different from 0.0) for most countries}
    \item{Direct energy input to economy ($q$) is important as a LINEX predictor of economic output ($y$)}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Tentative Conclusions} %----------------------------
  \begin{itemize}
  \item{Including energy in Cobb-Douglas and CES production functions is unwarranted \\
        \scriptsize
        Statistical evidence indicates that energy parameters are not significant for most countries \\Why? Energy is \emph{always already} embedded in capital stock ($K$) and labor ($L$), so including energy is unnecessary.
        \normalsize}
  \item{Including energy in LINEX production function is required \\
        \scriptsize
        The foundation of LINEX is an energy term.\\Statistical evidence indicates that LINEX parameters $a_0$ and $c_t$ are significant for most countries.\\LINEX fits overall are worse than Cobb-Douglas or CES.
        \normalsize}
  \vspace{\baselineskip}
  \pause

  \item{Need a modeling method (that is better than LINEX) that captures energy's embeddedness \\
        \scriptsize
        System dynamics modeling?
        \normalsize}
  \end{itemize}
\end{frame}

>>>>

% The equation for a linex model
% \newcommand{\linexEqn}{$y = e \: \mathrm{e}^{\left[a_0 \left(2 - \frac{l + e}{k} \right) + a_0 c_t \left(\frac{l}{e} - 1 \right)\right]}$}

\newcommand{\linexEqn}{$y = \mathrm{e}^{\left[a_0 \left(2 - \frac{l + e}{k} \right) + a_0 c_t \left(\frac{l}{e} - 1 \right)\right]} \times e$}

@@

\begin{frame}{Oil and Recessions}{\citep{Hamilton:2011fo}} %----------------------------
  \begin{center}
    10 of last 11 US recessions preceeded by oil price shocks
    \vspace{\baselineskip}

    \pause
    \small Exception: mild recession of 1960-1961 \normalsize
  \end{center}
\end{frame}

@@

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
    \pause
    \vspace{\baselineskip}

    \footnotesize
    No known rigorous, statistical comparative studies of economic growth models for various types of economies. 
    \normalsize
    \pause
    \vspace{\baselineskip}

    Extra credit: reflect dynamics of energy cost share and prices spikes.
  \end{center}
\end{frame}

@@

\begin{frame}{Economic Definitions} %----------------------------
  \begin{itemize}
    \item{Output ($Y$): total value (\$) of goods produced (GDP)}
    \item{Capital Stock ($K$): market value of productive goods (machines, tools, infrastructure).}
    \item{Labor ($L$): workers in an economy (people or hours)}
  \end{itemize}
\end{frame}

@@

\minorSeparator{Energy Data} %----------------------------

\begin{frame}{Energy Definitions} %----------------------------
  \begin{itemize}
    \item{Thermal energy ($Q$): heat content of an energy carrier}
    \item{Exergy ($X$): maximum work from a heat carrier}
    \item{Useful work ($U$): energy that generates economic activity directly}
  \end{itemize}
  \begin{center}Note: indexed $Q$ and $X$ are identical unless energy mix changes significantly. \end{center}
\end{frame}

@@

\begin{frame}{Sources of Energy Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{IEA} & \multirow{3}{*}{\includegraphics[scale=0.20]{figure_other/Figure_IEA.png}} & Coal, Oil, Petroleum, & \multirow{3}{*}{GJ/year} \\
                              &                          & Nat. gas, Nuclear, & \\
                              &                          & Renewables & \\
      \hline
      \multirow{4}{*}{FAOStat} & \multirow{4}{*}{\includegraphics[scale=0.07]{figure_other/Figure_FAOStat.png}} & Fuel wood,    & \multirow{4}{*}{GJ/year} \\
                              &                          & Available livestock &                                \\
                              &                          & (1980-2011: Developed; &                         \\
                              &                          & 1991-2011: Other) &                         \\
      \hline
      \multirow{2}{*}{REXS} & \multirow{2}{*}{\includegraphics[scale=0.05]{figure_other/Figure_Warr.png}} & Energy allocation factors. &  \multirow{2}{*}{\citet{Warr:2006hc}}\\
                              &                          & (1980-2000: Developed)     & \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

@@

    \begin{itemize}
      \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
      \begin{itemize}
        \item{Represents the irrepressible ``progress'' of technology}
        \item{Is independent of factor of production ($f$)}
      \end{itemize}
      \item{Used for illustration purposes here}
      \item{Example: \citet{Nel:2009wq} use $f$ = $q$}
      \item{$\lambda$ and $m$ estimated by \texttt{nls} function in \texttt{R}}
    \end{itemize}
\end{frame}

@@

\begin{frame}{Parameter Estimation for Cobb-Douglas With Energy} %----------------------------
  \begin{center}\cdeEqn\end{center}
  \begin{itemize}
    \item{Objective: determine $\lambda$, $\alpha$, $\beta$, and $\gamma$ that best fits historical data}
    \item{Perform grid search over $\gamma$: Re-estimate $\lambda$, $\alpha$, and $\beta$ at each value of $\gamma$}
    \item{Respect sum constraint: $\alpha + \beta + \gamma = 1$}
    \item{Respect absolute constraints: $0 \leq \alpha, \beta, \gamma \leq 1$} (not yet)
    \item{$\lambda$, $\alpha$, $\beta$, and $\gamma$ estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}

@@

    \item{Are we sure? (Is energy statistically significant?)\\Yes, if $CI_{\gamma}$ (95\%) doesn't cross 0.0.}
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item{Energy parameter ($\gamma$) has either}
      \begin{itemize}
        \item{Small magnitude (US, JP, CN, SA, IR)}
        \item{Large CI (ZA, TZ, ZM)}
        \item{Exception: UK}
      \end{itemize}
    \vspace{\baselineskip}

@@

\begin{frame}{Comments on CES Model Equation} %----------------------------
  \begin{center} \small \cesEqn \normalsize \end{center}
  \begin{itemize}
    \item{$\mathrm{e}^{\lambda(t-t_0)}$: Time-dependent technology term}
    \item{$\delta$: split between ($kl$) and ($e$)}
    \item{$\sigma \equiv \frac{1}{1+\rho}$: elasticity of substitution ($0 \leq \sigma \leq 1$)}
    \begin{itemize}
      \item{$\sigma \rightarrow 0$: ($kl$) and ($e$) are complementary}
      \item{$\sigma \rightarrow 1$: ($kl$) and ($e$) are substitutes}
    \end{itemize}
    \item{Example: \citet{Stern:2012ey}}
    \item{$\gamma$, $\lambda$, $\delta_1$, $\rho_1$, $\delta$, $\rho$ estimated by \texttt{cesEst} function in \texttt{R}'s \texttt{micEcon} package}
  \end{itemize}
\end{frame}

@@

    \item{Are we sure? (Is energy statistically significant?)\\Yes, if $CI_{\delta}$ (95\%) doesn't cross 0.0.}
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item{Energy parameter ($\delta$) has either}
      \begin{itemize}
        \item{Large magnitude (US, UK, JP, SA, IR)}
        \item{Large CI (CN, TZ, ZA, ZM)}
      \end{itemize}
      \vspace{\baselineskip}          
      \pause

    \item{Direct energy input to economy ($q$) is unimportant as a CES predictor of economic output ($y$)}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Comments on LINEX Model Equation} %----------------------------
  \begin{center} \linexEqn \end{center}
  \begin{itemize}
    \item{Derived from thermodynamics \citep{Kummel:1980wx}}
    \item{Time-dependent technology $\rightarrow$ efficiency ratios}
    \item{Productivity $\uparrow$ if}
    \begin{itemize}
      \item{Efficiency of producing capital stock increases ($k$ > $l + e$)}
      \item{Energy efficiency of labor increases ($l$ > $e$)}
    \end{itemize}
    \item{Example: \citet{Warr:2006hc}}
    \item{$a_0$ and $c_t$ estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}

@@

    \item{Are we sure?\\Yes, by definition!}
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item{LINEX parameters $a_0$ and $c_t$ are significant (different from 0) for most countries}
    \item{Direct energy input to economy ($q$) is important as a LINEX predictor of economic output ($y$)}
  \end{itemize}
\end{frame}

@@

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
  \end{center}
\end{frame}

\begin{frame}{Tentative Conclusions} %----------------------------
  \begin{itemize}
  \item{Including energy in Cobb-Douglas and CES production functions is unwarranted \\
        \scriptsize
        Preliminary statistical evidence indicates that energy parameters are not significant for most countries \\Why? Energy is \emph{always already} embedded in capital stock ($K$) and labor ($L$), so including energy is unnecessary.
        \normalsize}
  \item{Including energy in LINEX production function is required \\
        \scriptsize
        The foundation of LINEX is an energy term.\\Statistical evidence indicates that LINEX parameters $a_0$ and $c_t$ are significant for most countries.\\However, LINEX fits are worse overall than Cobb-Douglas or CES.
        \normalsize}
  \end{itemize}
\end{frame}

\begin{frame}{The ``Answer''} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
    \vspace{\baselineskip}

    \pause
    Answer: Cobb-Douglas without energy or CES 
  \end{center}
\end{frame}

\begin{frame}{Extra Credit?} %----------------------------
  \begin{itemize}
    \item{Price, cost share exogenous to all models}
    \pause
    \item{Models (CD without energy, CES) could be \emph{included within} a model that endogenizes price and cost share}
    \pause
    \item{Still searching for a modeling approach that captures price and cost share dynamics \\
      \scriptsize System dynamics modeling? \normalsize}
    \pause
    \item{No extra credit!}
  \end{itemize}
\end{frame}
